The Galactic Midplane Is Not a Plane: Implications for Dynamical Analysis with \textit{Gaia} Data and Beyond

### 2019-05-17 paper/ms.tex
Small changes throughout

This example uses an offset in $z$, but one 
%mm would expect an analogous effect from an offset in another coordinate, 
    analagous effects occur from offsets in other coordinates, 
such as the distance to the Galactic
center or the LSR. A similar effect in which actions gain time-dependence due to
a time-varying potential was pointed out by \citet{2015A&A...584A.120B}.

@@

Under this approximation, we can write down the cylindrical components
of the orbits as
\beq\label{eq:orbits_epi}
\begin{split}
\phi(t) &= \Omega_c t \\
R(t) &= R_g + A_R \sin{(\kappa t + \alpha)} \\
z(t) &= A_z \sin{(\nu t + \beta)}
\text{,}
\end{split}
\eeq
where $\kappa$ and $\nu$ are the radial/epicyclic and vertical frequencies, $\Omega_c
\equiv v_c/R_g$ is the orbital frequency of the guiding center, $A_z$ and $A_R$ are the amplitudes in the vertical and radial coordinates, and $\alpha$ and
$\beta$ are 
%mm arbitrary phases
    the initial orbital phases at $t = 0$.
Similarly, the velocities of the orbit are given by:
\beq\label{eq:orbits_vel_epi}
\begin{split}
v_{\phi}(t) &= v_c \\
v_R(t) &= \kappa A_R \cos{(\kappa t + \alpha)} \\
v_z(t) &= \nu A_z \cos{(\nu t + \beta)}
\text{.}
\end{split}
\eeq

@@

Gaussian summary statistics are clearly insufficient to describe the distribution shown in
Figure~\ref{fig:Jz_hist}. We therefore elect to measure this error by computing one
half the 95th percentile minus the 5th percentile of the distribution
of action values. We plot this quantity
%mm
     in Figure~\ref{fig:Jz_hist}, 
which we refer to as $\Delta J_i$ for each action, 
%mm in Figure~\ref{fig:Jz_hist} 
as a horizontal arrow anchored on the true action value. Because of the bimodality
of the error distribution, this quantity roughly measures the distance from
the true action value to the peak of one of the modes. Furthermore, this
bimodality also implies that $\Delta J_i$ is not very sensitive to the exact
percentiles used. This summary statistic does not reflect the bias induced
when the midplane error is~$\gtrsim z_{\text{max}}$.

@@

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{fig/schmactions_many_orbits_Jz_fun.pdf}
\end{center}
\caption{The fractional error in $J_z$ as a function of $J_z$ for a few
different offsets in $z$. All orbits have the same initial position of $(8, 0,
0)\,\kpc$ and velocity $(0, -190, v_z)\,\kms$, where we vary $v_z$.\footref{note:vz_orbits} We show
this for a $z$ offset of $10, 50, 100\,\pc$ (orange, teal, and green,
respectively). As before, the error ($\Delta J_z$) is one half the 95th
minus 5th percentile of the distribution of $J_z$ values over the
course of the orbit. There are large errors for \thin{}-like orbits ($J_z
\sim 0.5\,\actunit$), even for a small midplane offset of $10\,\pc$. As dashed
lines in each color we also plot the prediction for $\Delta J_z/J_z$ from the
epicyclic approximation (Equation~\eqref{eq:Ji_err_mosttime}), which shows
excellent agreement with the numerically computed values.}
\label{fig:dJz_fun_Jz}
\end{figure}

To further understand the effect of the midplane error, we also plot the
fractional error in $J_z$ as a function of $J_z$ for $z$ offsets of 
%mm $10, 50,
     10, 50, and $
100\,\pc$ (orange, teal, and green, respectively) in
Figure~\ref{fig:dJz_fun_Jz}. 
For each orbit, we set the initial position to be
$(8,0,0)\,\kpc$ and the initial velocity to be $(0, -190, v_z)\,\kms$, where
we vary $v_z$.\footnote{\label{note:vz_orbits}One can recover \thin{},
  \thick{}, and \halo{} by setting $v_z=10$, $50$, and $190
%mm
     \kms$,
 respectively, 
%mm [one significant figure values]
      giving $J_z \simeq 0.7, 20,$ and $500\,\actunit$  
(see Table~\ref{tab:orbits}).} For a \thin{}-like orbit ($J_z\sim
%mm 0.5
         0.7%
\,\actunit$), 
even a
$10\,\pc$ offset in $z$ is enough to induce a 
%mm $\sim25\%$ 
         $\sim20\%$ 
error in $J_z$. For
larger values of $J_z$, the fractional errors are suppressed, but the induced
error can still be large depending on how great the $z$ offset is. We also
plot the epicylic prediction for $\Delta J_z / J_z$ from
Equation~\eqref{eq:Ji_err_mosttime} as dashed lines for each $z$ offset. We
find that the epicyclic approximation matches the numerical estimate quite
well.

\section{
%mm Coordinate System Offsets from Azimuthal 
Midplane Variations}
\label{sec:local_fire}

@@

For this work,
we focus on the three Milky Way-mass zoom-ins considered in
\citet{2018arXiv180610564S}, which were simulated as part of the
\textit{Latte} suite and show broad agreement of many of their global
properties with observations of the Milky Way \citep{2016ApJ...827L..23W,
2018MNRAS.481.4133G}. The 
%mm $\z = 0$ snapshots\footnote{In this work, to avoid
%confusion with the vertical height $z$, we refer to cosmological redshift as
%$\z$.} of these three simulations, 
      The snapshots of these three simulations at cosmological redshift $\z = 0$,
named \mi{}, \mf{}, and \mm{},
%mm for shorthand, 
are publicly available alongside associated mock \textit{Gaia} DR2
catalogues generated from them.\footnote{\url{http://ananke.hub.yt}}

@@

The mass resolution of the simulation is 
%mm $2.6\times10^4$, $1.2\times10^4$, 
       $2.6\times10^4\,\Msun$, $1.2\times10^4\,\Msun$, 
and $1.0\times10^4\,\Msun$ for the dark matter, disk, and bulge components,
respectively. For the disk and bulge components, a softening length of
$30\,\pc$ is used whereas for the halo a softening length of $60\,\pc$ is
used. For Sagittarius, the softening length for the dark matter and
the stars is 
%mm $60$
     $60\,\pc$ 
and $40\,\pc$, respectively.

@@

\subsection{The Local Midplane} \label{ssec:local_midplane}
Using the two sets of simulations, we determine the local midplane
%mm 
       as a function of azimuth at the solar circle
that an observer might measure if they were situated in each of these
galaxies% 
%mm as a function of azimuth at the solar circle
. Starting from the coordinate system
described in the previous section, which is 
approximately 
%mm
    \MM{[isn't this by definition, at least on average, not
      approximate?]} 
aligned so that the
$z$-coordinate is perpendicular to the disk plane, we place our imaginary
observer at $z=0$ and a galactocentric cylindrical radius of $8.2\,\kpc$ and
vary the azimuth between $0<\phi<2\pi$. 
%mm [this was written backwards.  Reordered.] 
%We consider $50$ equally spaced bins in azimuth, a sufficient number
%such that each cylinder shares no stars with its neighboring
%cylinder. 
At 
%mm each value of 
    equally spaced values of 
$\phi$ we then compute the median $z$ for stars within a cylinder of radius $0.5\,\kpc$ and
height $1\,\kpc$ perpendicular to the fiducial disk  
%mm .
     and centered on it. We choose to use 50 values, sufficiently few that no cylinder shares stars
     with its neighbors.
We then re-define the new
midplane of the cylinder to be the median $z$, re-select stars, and iterate
until the median $z$ value converges. We find that only $10$ iterations of
this procedure are necessary for convergence. The resulting median $z$ is
taken to be what our observer would measure as the local Galactic midplane at
each $\phi$.

@@

To allow for potential small inaccuracies in the determination of the original
fiducial coordinate system, we also subtract the best fit 
%mm \text{A} \sin{\left(\phi + \text{B}\right)} + \text{C}\eeq curve
    curve of the form 
    \beq A \sin{\left(\phi + B\right)} + C \label{eq:fit-curve}\eeq 
from the midplane as a
function of azimuth to account for an overall tilt of the midplane (a
simplified version of the strategy described in
\citealt{2019ApJ...871..145A}). 
%mm [perhaps this would be better presented as a table] For
%$\text{A}$, the values are $-165$, $45$, and $8.8\,\pc$, for
%$\text{B}$ the values are $0.67$, $-0.088$, and $0.031\,\text{rad}$
%and for $\text{C}$ the values are $-69$, $19$, and $-18\,\pc$ for
%\mi{}, \mf{}, and \mm{}, respectively. 
    For models (\mi{}, \mf{}, \mm{}) the best fit values are $A =
    (-165, 45, 8.8)\,\pc$, $B = (0.67, -0.088, 0.031)\,\text{rad}$,
    and $C = (-69, 19, -18)\,\pc$.
For the assumed Solar radius of $8.2\,\kpc$, we can approximate the
angle offset 
%mm 
    $\Delta \theta$ 
for the $z$-axis from the values of 
%mm [change to rad for consistency with B.   Note that you did not
%give two significant figures of accuracy for the last value.
%Strictly speaking should be corrected, though this is pretty minor.]
%$\text{A}$ --- we compute $1.15$, $0.31$, and $0.06\,\deg$ for \mi{},
%\mf{}, and \mm{}.  
      $A$. We find for the same models $\Delta \theta = (0.020, 5.4
      \times 10^{-3}, 1.0 \times 10^{-3})\,\text{rad}$.
These angle offsets are consistent with the
values given in \citet{2018arXiv180610564S} for the difference between the
$z$-axis as defined by the gas and stars.

@@

\acknowledgments
We would like to thank Megan Bedell, Robert A. Benjamin, Tobias Buck, Elena
D'Onghia, Benoit Famaey, 
%mm and Adrian Price-Whelan
    Adrian Price-Whelan, and Todd Phillips
for helpful
discussions. 
%mm A.B. would like to thank Todd Phillips for helpful discussions. 
A.B. was supported
in part by the Roy \& Diana Vagelos Program in the Molecular Life Sciences and
the Roy \& Diana Vagelos Challenge Award. K.V.J.'s, C.F.P.L.'s, and R.E.S.'s work was
performed in part during the Gaia19 workshop and the 2019 Santa Barbara Gaia
Sprint (also supported by the Heising-Simons Foundation), both hosted by the
Kavli Institute for Theoretical Physics at the University of California, Santa
Barbara. This work used the Extreme Science and Engineering Discovery
Environment (XSEDE), 
%mm
    \MM{[Does anyone want their XSEDE grant number here?]}
which is supported by 
%mm National Science Foundation 
     NSF
grant 
%mm number 
OCI-1053575. K.V.J.'s contributions were supported in part by the
%mm National Science Foundation 
     NSF
under grants 
%mm NSF 
PHY-1748958 and 
%mm NSF 
AST-1614743.
M.-M.M.L. was partly supported by NSF grant 
%mm AST18-15461. [for consistency with above]
        AST-1815461. 
%mm The work of D.A.-A., A.B., D.W.H., R.E.S., K.V.J., M.-M.M.L., and
%M.K.N. is supported by the Simons Foundation. [This is redundant with
%our Flatiron Institute affiliations, unless there is additional,
%separate, support for any of us.]

>>>>

This example uses an offset in $z$, but 
%mm would expect an analogous effect from an offset in another coordinate, 
    analagous effects occur from offsets in other coordinates, 
such as the distance to the Galactic
center or the LSR. A similar effect in which actions gain time-dependence due to
a time-varying potential was pointed out by \citet{2015A&A...584A.120B}.

@@

Under this approximation, we can write down the cylindrical components
of the orbits as
\beq\label{eq:orbits_epi}
\begin{split}
\phi(t) &= \Omega_c t \\
R(t) &= R_g + A_R \sin{(\kappa t + \alpha)} \\
z(t) &= A_z \sin{(\nu t + \beta)}
\text{,}
\end{split}
\eeq
where $\kappa$ and $\nu$ are the radial/epicyclic and vertical frequencies, $\Omega_c
\equiv v_c/R_g$ is the orbital frequency of the guiding center, $A_z$ and $A_R$ are the amplitudes in the vertical and radial coordinates, and $\alpha$ and
$\beta$ are 
%mm arbitrary phases
    the initial orbital phases.
Similarly, the velocities of the orbit are given by:
\beq\label{eq:orbits_vel_epi}
\begin{split}
v_{\phi}(t) &= v_c \\
v_R(t) &= \kappa A_R \cos{(\kappa t + \alpha)} \\
v_z(t) &= \nu A_z \cos{(\nu t + \beta)}
\text{.}
\end{split}
\eeq

@@

Gaussian summary statistics are clearly insufficient to describe the distribution shown in
Figure~\ref{fig:Jz_hist}. We therefore elect to measure this error by computing one
half the 95th percentile minus the 5th percentile of the distribution
of action values. We refer to this quantity as $\Delta J_i$ for each action and plot it in Figure~\ref{fig:Jz_hist} as a horizontal arrow anchored on the true action value.
 Because of the bimodality
of the error distribution, this quantity roughly measures the distance from
the true action value to the peak of one of the modes. Furthermore, this
bimodality also implies that $\Delta J_i$ is not very sensitive to the exact
percentiles used. This summary statistic does not reflect the bias induced
when the midplane error is~$\gtrsim z_{\text{max}}$.

@@

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{fig/schmactions_many_orbits_Jz_fun.pdf}
\end{center}
\caption{The fractional error in $J_z$ as a function of $J_z$ for a few
different offsets in $z$. All orbits have the same initial position of $(8, 0,
0)\,\kpc$ and velocity $(0, -190, v_z)\,\kms$, where we vary $v_z$.\footref{note:vz_orbits} We show
this for a $z$ offset of $10$, $50$, and $100\,\pc$ (orange, teal, and green,
respectively). As before, the error ($\Delta J_z$) is one half the 95th
minus 5th percentile of the distribution of $J_z$ values over the
course of the orbit. There are large errors for \thin{}-like orbits ($J_z
\sim 0.5\,\actunit$), even for a small midplane offset of $10\,\pc$. As dashed
lines in each color we also plot the prediction for $\Delta J_z/J_z$ from the
epicyclic approximation (Equation~\eqref{eq:Ji_err_mosttime}), which shows
excellent agreement with the numerically computed values.}
\label{fig:dJz_fun_Jz}
\end{figure}

To further understand the effect of the midplane error, we also plot the
fractional error in $J_z$ as a function of $J_z$ for $z$ offsets of 
%mm $10, 50,
     $10$, $50$, and $
100\,\pc$ (orange, teal, and green, respectively) in
Figure~\ref{fig:dJz_fun_Jz}. 
For each orbit, we set the initial position to be
$(8,0,0)\,\kpc$ and the initial velocity to be $(0, -190, v_z)\,\kms$, where
we vary $v_z$.\footnote{\label{note:vz_orbits}One can recover \thin{},
  \thick{}, and \halo{} by setting $v_z=10$, $50$, and $190$,
 respectively, 
%mm [one significant figure values]
      giving $J_z \simeq 0.7, 20,$ and $500\,\actunit$  
(see Table~\ref{tab:orbits}).} For a \thin{}-like orbit ($J_z\sim
%mm 0.5
         0.7%
\,\actunit$), 
even a
$10\,\pc$ offset in $z$ is enough to induce a 
%mm $\sim25\%$ 
         $\sim20\%$ 
error in $J_z$. For
larger values of $J_z$, the fractional errors are suppressed, but the induced
error can still be large depending on how great the $z$ offset is. We also
plot the epicylic prediction for $\Delta J_z / J_z$ from
Equation~\eqref{eq:Ji_err_mosttime} as dashed lines for each $z$ offset. We
find that the epicyclic approximation matches the numerical estimate quite
well.

\section{
%mm Coordinate System Offsets from Azimuthal 
Azimuthal Midplane Variations}
\label{sec:local_fire}

@@

For this work,
we focus on the three Milky Way-mass zoom-ins considered in
\citet{2018arXiv180610564S}, which were simulated as part of the
\textit{Latte} suite and show broad agreement of many of their global
properties with observations of the Milky Way \citep{2016ApJ...827L..23W,
2018MNRAS.481.4133G}. The 
$\z = 0$ snapshots\footnote{In this work, to avoid
confusion with the vertical height $z$, we refer to cosmological redshift as
$\z$.} of these three simulations, 
      The snapshots of these three simulations at cosmological redshift $\z = 0$,
named \mi{}, \mf{}, and \mm{},
%mm for shorthand, 
are publicly available alongside associated mock \textit{Gaia} DR2
catalogues generated from them.\footnote{\url{http://ananke.hub.yt}}

@@

The mass resolution of the simulation is 
%mm $2.6\times10^4$, $1.2\times10^4$, 
       $2.6\times10^4$, $1.2\times10^4$, 
and $1.0\times10^4\,\Msun$ for the dark matter, disk, and bulge components,
respectively. For the disk and bulge components, a softening length of
$30\,\pc$ is used whereas for the halo a softening length of $60\,\pc$ is
used. For Sagittarius, the softening length for the dark matter and
the stars is 
     $60$ 
and $40\,\pc$, respectively.

@@

\subsection{The Local Midplane} \label{ssec:local_midplane}
Using the two sets of simulations, we determine the local midplane
%mm 
       as a function of azimuth at the solar circle
that an observer might measure if they were situated in each of these
galaxies% 
%mm as a function of azimuth at the solar circle
. Starting from the coordinate system
described in the previous section, which is 
approximately 
aligned so that the
$z$-coordinate is perpendicular to the disk plane at the solar circle, we place our imaginary
observer at $z=0$ and a galactocentric cylindrical radius of $8.2\,\kpc$ and
vary the azimuth between $0<\phi<2\pi$. 
%mm [this was written backwards.  Reordered.] 
%We consider $50$ equally spaced bins in azimuth, a sufficient number
%such that each cylinder shares no stars with its neighboring
%cylinder. 
At 
%mm each value of 
    equally spaced values of 
$\phi$ we then compute the median $z$ for stars within a cylinder of radius $0.5\,\kpc$ and
height $1\,\kpc$ perpendicular to the fiducial disk  
%mm .
     and centered on it. We choose to use 50 values, sufficiently few that no cylinder shares stars
     with its neighbors.
We then re-define the new
midplane of the cylinder to be the median $z$, re-select stars, and iterate
until the median $z$ value converges. We find that only $10$ iterations of
this procedure are necessary for convergence. The resulting median $z$ is
taken to be what our observer would measure as the local Galactic midplane at
each $\phi$.

@@

To allow for potential small inaccuracies in the determination of the original
fiducial coordinate system, we also subtract the best fit 
%mm \text{A} \sin{\left(\phi + \text{B}\right)} + \text{C}\eeq curve
    curve of the form 
    \beq A \sin{\left(\phi + B\right)} + C \label{eq:fit-curve}\eeq 
from the midplane as a
function of azimuth to account for an overall tilt of the midplane (a
simplified version of the strategy described in
\citealt{2019ApJ...871..145A}). 
%mm [perhaps this would be better presented as a table] For
%$\text{A}$, the values are $-165$, $45$, and $8.8\,\pc$, for
%$\text{B}$ the values are $0.67$, $-0.088$, and $0.031\,\text{rad}$
%and for $\text{C}$ the values are $-69$, $19$, and $-18\,\pc$ for
%\mi{}, \mf{}, and \mm{}, respectively. 
    For models (\mi{}, \mf{}, \mm{}) the best fit values are $A =
    (-165, 45, 8.8)\,\pc$, $B = (0.67, -0.088, 0.031)\,\text{rad}$,
    and $C = (-69, 19, -18)\,\pc$.
For the assumed Solar radius of $8.2\,\kpc$, we can approximate the
angle offset 
%mm 
    $\Delta \theta$ 
for the $z$-axis from the values of 
%mm [change to rad for consistency with B.   Note that you did not
%give two significant figures of accuracy for the last value.
%Strictly speaking should be corrected, though this is pretty minor.]
%$\text{A}$ --- we compute $1.15$, $0.31$, and $0.06\,\deg$ for \mi{},
%\mf{}, and \mm{}.  
      $A$. We find for the same models $\Delta \theta = (1.15, 0.31,
      0.062)\,\text{deg}$.
These angle offsets are consistent with the
values given in \citet{2018arXiv180610564S} for the difference between the
$z$-axis as defined by the gas and stars.

@@

\acknowledgments
We would like to thank Megan Bedell, Robert A. Benjamin, Tobias Buck, Elena
D'Onghia, Benoit Famaey, 
and Adrian Price-Whelan
    % Adrian Price-Whelan, and Todd Phillips
for helpful
discussions. 
A.B. would like to thank Todd Phillips for helpful discussions. 
A.B. was supported
in part by the Roy \& Diana Vagelos Program in the Molecular Life Sciences and
the Roy \& Diana Vagelos Challenge Award. K.V.J.'s, C.F.P.L.'s, and R.E.S.'s work was
performed in part during the Gaia19 workshop and the 2019 Santa Barbara Gaia
Sprint (also supported by the Heising-Simons Foundation), both hosted by the
Kavli Institute for Theoretical Physics at the University of California, Santa
Barbara. This work used the Extreme Science and Engineering Discovery
Environment (XSEDE), 
%mm
    \MM{[Does anyone want their XSEDE grant number here?]}
which is supported by 
%mm National Science Foundation 
     NSF
grant 
%mm number 
OCI-1053575. K.V.J.'s contributions were supported in part by the
%mm National Science Foundation 
     NSF
under grants 
%mm NSF 
PHY-1748958 and 
%mm NSF 
AST-1614743.
M.-M.M.L. was partly supported by NSF grant 
%mm AST18-15461. [for consistency with above]
        AST-1815461. 
%mm The work of D.A.-A., A.B., D.W.H., R.E.S., K.V.J., M.-M.M.L., and
%M.K.N. is supported by the Simons Foundation. [This is redundant with
%our Flatiron Institute affiliations, unless there is additional,
%separate, support for any of us.]
